GEEPERs: Principal Stratification using Principal Scores and Stacked Estimating Equations

### 2025-01-09 writeUps/paper.tex
paper edits

\documentclass[]{article}

@@

\usepackage[longnamesfirst,sort]{natbib}% Citation support using natbib.sty
\bibpunct[, ]{(}{)}{;}{a}{,}{,}% Citation support using natbib.sty
\newcommand\bibfont{\fontsize{10}{12}\selectfont}% To set the list of references in 10 point font using natbib.sty

@@

\begin{abstract}
Principal stratification is a framework for making sense of causal effects conditioned on variables that themselves may have been affected by treatment. For instance, one component of an educational computer application is the availability of “bottom-out” hints that provide the answer. In evaluating a recent experimental evaluation against alternative programs without bottom-out hints, researchers may be interested in estimating separate average treatment effects for students who, if given the opportunity, would request bottom-out hints frequently, and for students who would not. Most principal stratification estimators rely on strong structural or modeling assumptions, and many require advanced statistical training to fit and check. In this paper, we introduce a new M-estimation principal effect estimator for one-way noncompliance based on a binary indicator. Estimates may be computed using conventional regressions (though the standard errors require a specialized sandwich formula) and do not rely on distributional assumptions. We present a simulation study that shows that the novel method is more robust than popular alternatives and illustrate the method in an analysis of data on bottom-out hint requests.
\end{abstract}
% \begin{keywords}
% Causal Inference; Principal Stratification; Educational Technology
% \end{keywords}
%\end{frontmatter}

@@

%Randomized experiments easily admit subgroup analysis---estimation of treatment effects for a subset of the study population---when those subgroups are defined at baseline, and not themselves a function of treatment assignment.
%A simple, popular approach to estimating subgroup effects is to use ordinary least squares (OLS) regression, regressing the outcome on a treatment indicator interacted with an indicator for subgroup membership.
%However, when subgroup membership is a function of a post-treatment variable, then it may itself be affected by the treatment.
%Contrasts between treatment and control conditions within those subgroups may not have a causal interpretation at all---and when they do, the same techniques that work for subgroups defined at baseline %(such as OLS with interactions)
%will often produce inconsistent estimates.

%%% open with hook for ed researchers

@@

\citet{frangakis} introduced the terminology of principal stratification and extended the framework well beyond noncompliance, inspiring a broad literature introducing (and debating) novel applications and estimators.
For instance, \citet{fellerEtAl2016} estimated differential effects of random assignment to the Head Start early childhood care program for children who, if assigned to control, would receive home-based care, and for those who would attend a different early childhood care program.
Other methodologists have developed estimators for the evaluation of surrogate outcomes \citep{li2010bayesian}, the investigation of causal mechanisms \citep{lidsayPage}, sample attrition \citep{zhangRubin, ding2011}, and other problems.
Since interest is often in comparing effects between principal strata---a goal that runs counter to the exclusion restriction---the earliest principal stratification estimators built on \citet{imbens1997}'s parametric mixture models \citep[e.g.][]{Barnard01062003,mealli2004analyzing}, and that approach remains popular \citep{lidsayPage,fellerEtaAl2016}.
However, these models are often hard to specify and hard or impossible to test.
Even when they are well-specified they can yield estimates that are severely biased and misleading \citep{griffin2008application,feller2016principal}.
The most common type of parametric model, the normal mixture model, is especially dubious for studying bottom-outers in \citet{impactPaper}'s RCT, whose outcome, the total score on a 10-item posttest, is not normally distributed.

These concerns have prompted the development of alternative approaches, including bounding \citep{bounding} and randomization inference \citep{nolen2011randomization}.
Of particular importance is the ``principal score,'' \citep{jo,dingLu,feller2017principal}, or the probability that a subject belongs to a particular principal stratum, as a function of pre-treatment covariates.
Under a strong, untestable assumption called ``principal ignorability,''  principal scores can be transformed into weights to compute form an unbiased estimate of principal effects.
Principal ignorability is a direct analogue of the strong ignorability assumption in causal observational studies \citep[e.g.][]{rosenbaum1986central}---that all covariates that predict both principal stratum membership and outcomes have been accounted for---in other words, the precise type of assumption that randomized trials are designed to avoid.
A separate set of results in \citet{jo2002}, \citet{ding2011}, \citet{jiangDing2021}, and others have shown that, under certain circumstances, researchers can use baseline covariates to identify and estimate principal effects without relying on principal ignorability.

@@

This paper focuses on a special case of principal stratification---one-way, binary non-compliance---that is, there are only two principal strata and principal stratum membership is observable in one treatment arm, but not in the other.
This scenario would occur if, say, subjects in the control arm of a study had no access to the treatment, and not all subjects assigned to the treatment arm received the full treatment.
While such a setup may represent the simplest and easiest principal stratification problem, we believe it is a common scenario.
For example, \textsc{geepers} could be used to estimate effects from an RCT for participants categorized as low- or high-implementers. It could also be used to assess more granular questions about effect heterogeneity, such as whether an intervention's impact is contingent upon specific behaviors displayed by either practitioners or students.

@@

Now, the estimating equations $\sum_i\tilde{\Psi}_i=\bm{0}$ are algebraically equivalent to the estimating equations for a particular OLS fit, after some transformations, the most important of which follows.
Let
\begin{equation}\label{eq:ri}
\ri=Z_iS_i+(1-Z_i)\ppi=\begin{cases}
\ppi &Z_i=0\\
\sti &Z_i=1
\end{cases}
\end{equation}
In general, $(1-Z_i)\ppi=(1-Z_i)\ri$ and $Z_i\sti=Z_i\ri$, allowing $\ri$ to take the place of $\ppi$ and $S_i$ throughout $\tilde{\Psi}_i$.

@@

% If principal scores were known exactly, the M-estimation structure of $\heff0$ and $\heff1$ implies that their sampling variance could be estimated by the standard sandwich estimator for OLS \citep{stefanskiBoos}.
% This will rarely be the case.
%However,
If the principal scores are modeled as \eqref{eq:prinScoreMod}, and parameters $\bm{\alpha}$ are estimated by M-estimation, with estimating equations $\sum_{i: Z_i=1}\Omega_i=\bm{0}$, then the estimating equations for the principal score model and principal effect estimation can be stacked \citep[c.f.][]{boosStefanskiBook}, as
\begin{equation}\label{eq:stacked}
  \sum_{i=1}^n \begin{pmatrix}
    Z_i\Omega_i(\bxsi,S_i;\bm{\alpha})\\
    \Psi_i\left(Y_i,Z_i,\ri(\bm{\alpha}); \bm{\beta}\right)\end{pmatrix}=\begin{pmatrix} \bm{0}\\\bm{0}\end{pmatrix}
\end{equation}
where $\ri\left(\bm{\alpha}\right)=S_iZ_i+\ppip{\bm{\alpha}}(1-Z_i)$ emphasizes $\ppi$'s dependence on $\bm{\alpha}$.
In practice, the principal score model \eqref{eq:prinScoreMod} could be fit first, yielding estimates $\bm{\hat{\alpha}}$, and model \eqref{eq:regression0} could be fit next, using $\ri=\ri\left(\bm{\hat{\alpha}}\right)$.
Together, the vector of estimates $[\bm{\hat{\alpha}},\bm{\hat{\beta}}]$ represents a zero of the stacked estimating equations \eqref{eq:stacked}.

@@

In any event, Assumption \ref{ass:rci} allows principal effects to be estimated with an OLS model including $\bxy$:

\begin{prop}{\textsc{geepers}}\label{prop:reg2}

@@

Potential and observed outcomes were generated as
\begin{equation}\label{eq:y-sim}
Y_i=\beta_1\sti+\beta_3Z_i\sti+(\gamma_1+\gamma_2\sti+\gamma_3Z_i)(x_{1i}+x_{2i})+\frac{1}{\sqrt{6}}x_{3i}+\epsilon_i%\\
\end{equation}
%   \begin{split}
%   \yci&=\beta_1\sti+(\gamma_1+\gamma_2\sti)(x_{1i}+x_{2i})+\frac{1}{\sqrt{6}}x_{3i}+\epsilon_i\\
%   \yti&=0.3\sti+(\gamma_1+\gamma_2\sti+\gamma_3Z_i)(x_{1i}+x_{2i})+\frac{1}{\sqrt{6}}x_{3i}+\epsilon_i\\
%   Y_i&=Z_i\yti+(1-Z_i)\yci
%   \end{split}
% \end{equation}
with $\beta_3=0.3-\beta_1$, $\EE[\epsilon_i]=0$, and $Var(\epsilon_i)=1/2$.
The distribution of $\epsilon_i$, $\beta_1$, and values of coefficients $\gamma_1$, $\gamma_2$, and $\gamma_3$ were manipulated factors in the study.
As described below, in most simulated scenarios, $\gamma_1=1/\sqrt{6}$ and $\gamma_2=\gamma_3=0$. In these cases there is no average treatment effect in the $\st=0$ stratum, and, since  $Var(\gamma_1(x_1+x_2+x_3))=1/2$ and $Var(\epsilon_i)=1/2$, covariates explained half of the variance of $Y$ within treatment group and principal stratum, and observed covariates $x_1$ and $x_2$ explained 1/3 of that variance.
These values are in line with estimates from outcome models in Section \ref{sec:fh26} and are typical in education contexts \cite{hedgesHedberg}.

@@

\subsection{Data Analysis}
We estimated principal scores using logistic regression of observed $S$ on school fixed effects and a set of baseline variables $\bxs$.
Three criteria governed our choice of covariates and regression specifications for estimating principal scores and effects.
First, Proposition \ref{prop:reg2} requires consistent estimation of principal scores and (via Assumption \ref{ass:rci}) correct specification of the outcome model, that is, the model for $\EE[Y|\bxy,Z,\st]$.
Hence, it is essential to choose covariates and transformations $\bxs$ and $\bxy$ that lead to good model fit.

Second, Assumption \ref{ass:vps} requires principal scores to vary with at least some element $\bxs$. Moreover, the results of the simulation study suggest that when $\st$ can be predicted more precisely as a function of $\bxs$, \textsc{geeper}s principal effect estimators are more precise and more robust to model misspecification.
Those considerations militate in favor of choosing covariates $\bxs$ to optimize out-of-sample principal score prediction accuracy.
There is good reason to believe that analogous arguments apply to selection of outcome covariates $\bxy$.

The first two criteria motivated our choice of two strategies for choosing $\bxs$ and $\bxy$.
In our main model, we included every available covariate in the dataset and chose transformations of those covariates based on the results of residual and binned residual plots \citep{arm}.
Including all possible covariates maximizes the model's opportunities to detect covariate relationships that would predict $Y$ or $\st$.
However, this strategy can also lead to overfitting, which would decrease out-of-sample predictive accuracy.
In an alternative analysis, we chose a subset of covariates to include in $\bxs$ and $\bxy$ with a backward stepwise selection algorithm minimizing the AIC \citep{aic} of the principal score model and then modified the resulting models based on residual plots.
The model selection process is recorded in the GitHub repository cited above.

Lastly, Assumption \ref{ass:ci} implies that there is no interaction between $\bxs$ and $\st$ in the outcome model. Further, for the sake of statistical efficiency, we did not consider outcome models with interactions between covariates and $Z$.
\emph{A priori}, the student attribute that poses the greatest threat to these modeling choices is prior achievement, that is, scores on the pretest and the fifth-grade state test.
This is both because they measure a similar construct to the posttest, and because \cite{impactPaper} found a pretest-treatment interaction in a different treatment contrast.
Following that reasoning, we conducted a third parallel analysis excluding prior achievement covariates in $\bxs$ but including all covariates in $\bxy$.

@@

Outcome models for all analyses adjusted for all of the variables that were in the corresponding principal score models, in addition to measurements of students' prior knowledge; also, the models substituted class fixed effects for school fixed effects.\footnote{Why use a coarser set of fixed effects for the principal score model than the outcome regression? First, the principal score model was fit to a smaller sample (only students in the ``Instant'' condition), so some classrooms only included one student, which can lead to poor fit. Second, the asymptotics of logistic regression are more restricted than OLS, so logistic regression tends to perform poorly when the number of parameters increases with the sample size \citep{agresti}.}
As with principal score models, we checked model fit with residual plots and modified model specifications accordingly.
%\subsubsection{Principal Score Model}\label{sec:psMod}

@@

\subsection{Results}

@@

% \subsubsection{Outcome Regression and Principal Effects}\label{sec:outMod}

% Figure \ref{fig:effects} plots the estimated principal effects for bottom-outers and non-bottom-outers, using two different models---alongside estimates of the overall average treatment effect, labeled ``ATE.''
% The two sets of estimated differed in that the first assumed that coefficients on covariates were equal across principal strata, whereas the second set of results allowed the coefficients on pretest and 5th-grade standardized test scores---the most important predictors of outcomes and principal strata membership, respectively---to vary between strata.

% The results suggest little difference, if any, between the effects in the two principal strata.
% Estimated effects compared to BAU or DragonBox were slightly lower for bottom-outers than for non-bottom-outers, and estimated effects compared to FH2T were higher for non-bottom-outers.
% However, none of these differences is statistically significant, so opposite patterns, or no difference between principal effects at all, would also be consistent with the data.
% In fact, 95\% confidence intervals for all principal effects include both positive and negative principal effects.
% There is no discernible difference between estimated principal effects when coefficients on pretest and grade-5 standardized tests were or were not allowed to vary between strata, though standard errors from the models including an interaction were somewhat higher.

\begin{figure}
  \centering
  \includegraphics{compareMethods.pdf}
  \caption{Principal effect estimates with 95\% confidence or credible intervals using three different principal stratification methods, alongside ATE estimates and 95\% confidence intervals, for contrasts between the ASSISTments condition and each of the three other conditions in the study. The \textsc{geepers} estimates used the ``All Covariates'' principal score model.}
  \label{fig:compare}
\end{figure}

Figure \ref{fig:compare} shows principal effect estimates from \textsc{geepers} (using the All Covariates principal score model) compared to analogous estimates using Bayesian mixture modeling and \textsc{psw}, alongside estimates of the overall average treatment effects, labeled ``ATE.''
The \textsc{psw} model used the same principal score model as \textsc{geepers}, while the Bayesian model used the same principal score and outcome model specifications.

@@

\sloppy
Following \eqref{eq:stacked}, let $\blam(Z_i,S_i,\bxsi,\bxy_i,Y_i;\bm{\alpha},\bm{\beta})=\begin{pmatrix} Z_i\Omega(\bxsi,S_i;\bm{\alpha})\\ \Psi(\bxy_i,\bxsi,Y_i,Z_i,\ri(\bm{\alpha});\bm{\beta})\end{pmatrix}$, the stacked estimating equations of \eqref{eq:psMod} and \eqref{eq:regression}.
Going forward, for the sake of brevity, we will write $\blam_i(\bm{\alpha},\bm{\beta})=\blam(Z_i,S_i,\bxsi,\bxy_i,Y_i;\bm{\hat{\alpha}},\bm{\hat{\beta}})$, where dependence on the data for $i$ is captured in the subscript $i$, with similar meanings for $\Psi_i(\bm{\alpha},\bm{\beta})$ and $\omega_i(\bm{\alpha})$.
%Then let $\blamh_i=\blam_i(Z_i,S_i,\bxsi,\bxy_i,Y_i;\bm{\hat{\alpha}},\bm{\hat{\beta}})$ be the estimating equations evaluated at the estimated parameters, and let
%\begin{equation*}
%\dblam_i=\frac{\partial}{\partial [\alpha,\beta]'} \blam_i\Bigr|_{\substack{\bm{\alpha}=\bm{\hat{\alpha}}\\\bm{\beta}=\bm{\hat{\beta}}}}
%\end{equation*}
%the derivative matrix of $\blam$ evaluated at the estimated parameters.

@@

Following \citet[][p. 373]{carroll2006measurement}, we %separate the estimating equations $\Psi$ into $\phi$ and $\psi$, estimating equations for models \eqref{
can decompose the matrices into diagonal elements
\begin{equation*}
    \begin{split}
        A_{1,1}&=\sum_i \partial \Omega_i/\partial \bm{\alpha}|_{\bm{\alpha}=\bm{\hat{\alpha}}}\\
        A_{2,2}&=\sum_i\partial\Psi_i/\partial \bm{\beta}|_{\bm{\beta}=\bm{\hat{\beta}}}\\
        B_{1,1}&=\sum_i\Omega_i(\bm{\hat\alpha})\Omega_i(\bm{\hat\alpha})'\\
        B_{2,2}&=\sum_i \Psi_i(\bm{\hat\alpha},\bm{\hat\beta})\Psi_i(\bm{\hat\alpha},\bm{\hat\beta})'
    \end{split}
\end{equation*}
 that pertain to the parameter sets $\bm{\alpha}$ and $\bm{\beta}$ and the estimating equations for models \eqref{eq:psMod} and  \eqref{eq:regression}, respectively, and
 \begin{equation*}
     \begin{split}
         A_{21}&=\sum_i\partial\Psi_i/\partial \bm{\alpha}|_{\bm{\alpha}=\bm{\hat{\alpha}}}\\
         B_{12}=B_{21}'&=\sum_i \Omega_i(\bm{\hat{\alpha}})\Psi_i(\bm{\hat{\alpha}},\bm{\hat{\beta}})'
     \end{split}
 \end{equation*}
 which capture the dependence of model \eqref{eq:regression} on the parameters $\bm{\alpha}$ from \eqref{eq:psMod} and the covariance between the estimating equations of the two models.
The sub-matrix $A_{12}=\sum_i \partial \Omega_i/\partial \bm{\beta}=0$, since \eqref{eq:psMod} does not depend on $\bm{\beta}$.

@@

\begin{center}
    \includegraphics[width=5.5in]{simFigs/alphaAUC.pdf}
\end{center}
%\begin{table}
%\caption{Coverage of nominal 95\% confidence intervals}
\clearpage

\subsection{Full Empirical 95\% Interval Coverage Results}
The following table gives the empirical coverage of nominal 95\% intervals for \textsc{geepers} and mixture model principal effect estimates under varying data generating models.\\
\input{writeUps/coverageTabAppendix.tex}

\clearpage
\subsection{Full RMSE Results}
The following table gives the root mean squared error (RMSE) for \textsc{geepers}, mixture model, and principal score weighting principal effect estimates under varying data generating models.\\
\input{writeUps/rmseTabAppendix}

@@

\small
\input{writeUps/tab1num}

@@

\small
%\input{writeUps/outcomeRegAppendix}
\input{writeUps/psTab}

\input{writeUps/regTab}

>>>>

\documentclass[]{article}
\usepackage[letterpaper]{geometry}

@@

\usepackage[longnamesfirst,sort]{natbib}% Citation support using natbib.sty
\bibpunct[, ]{(}{)}{;}{a}{,}{,}% Citation support using natbib.sty
\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% To set the list of references in 10 point font using natbib.sty

@@

\maketitle

\begin{abstract}
Principal stratification is a framework for making sense of causal effects conditioned on variables that themselves may have been affected by treatment. For instance, one component of an educational computer application is the availability of “bottom-out” hints that provide the answer. In evaluating a recent experimental evaluation against alternative programs without bottom-out hints, researchers may be interested in estimating separate average treatment effects for students who, if given the opportunity, would request bottom-out hints frequently, and for students who would not. Most principal stratification estimators rely on strong structural or modeling assumptions, and many require advanced statistical training to fit and check. In this paper, we introduce a new M-estimation principal effect estimator for one-way noncompliance based on a binary indicator. Estimates may be computed using conventional regressions (though the standard errors require a specialized sandwich formula) and do not rely on distributional assumptions. We present a simulation study that shows that the novel method is more robust than popular alternatives and illustrate the method in an analysis of data on bottom-out hint requests.
\end{abstract}
% \begin{keywords}
% Causal Inference; Principal Stratification; Educational Technology
% \end{keywords}

@@

@@

\citet{frangakis} introduced the terminology of principal stratification and extended the framework well beyond noncompliance, inspiring a broad literature introducing (and debating) novel applications and estimators.
For instance, \citet{fellerEtAl2016} estimated differential effects of random assignment to the Head Start early childhood care program for children who, if assigned to control, would receive home-based care, and for those who would attend a different early childhood care program.
Other methodologists have developed estimators for the evaluation of surrogate outcomes \citep{li2010bayesian}, the investigation of causal mechanisms \citep{lidsayPage}, sample attrition \citep{zhangRubin, ding2011}, and other problems.
Since interest is often in comparing effects between principal strata---a goal that runs counter to the exclusion restriction---the earliest principal stratification estimators built on \citet{imbens1997bayesian}'s parametric mixture models \citep[e.g.][]{Barnard01062003,mealli2004analyzing}, and that approach remains popular \citep{lidsayPage,fellerEtAl2016}.
However, these models are often hard to specify and hard or impossible to test.
Even when they are well-specified they can yield estimates that are severely biased and misleading \citep{griffin2008application,feller2016principal}.
The most common type of parametric model, the normal mixture model, is especially dubious for studying bottom-outers in \citet{impactPaper}'s RCT, whose outcome, the total score on a 10-item posttest, is not normally distributed.

These concerns have prompted the development of alternative approaches, including bounding \citep{bounding} and randomization inference \citep{nolen2011randomization}.
Of particular importance is the ``principal score,'' \citep{jo,dingLu,feller2017principal}, or the probability that a subject belongs to a particular principal stratum, as a function of pre-treatment covariates.
Under a strong, untestable assumption called ``principal ignorability,''  principal scores can be transformed into weights to compute form an unbiased estimate of principal effects.
Principal ignorability is a direct analogue of the strong ignorability assumption in causal observational studies \citep[e.g.][]{rosenbaum1983central}---that all covariates that predict both principal stratum membership and outcomes have been accounted for---in other words, the precise type of assumption that randomized trials are designed to avoid.
A separate set of results in \citet{jo2002}, \citet{ding2011}, \citet{jiangDing2021}, and others have shown that, under certain circumstances, researchers can use baseline covariates to identify and estimate principal effects without relying on principal ignorability.

@@

@@

Now, the estimating equations $\sum_i\tilde{\Psi}_i=\bm{0}$ are algebraically equivalent to the estimating equations for a particular OLS fit, after some transformations, the most important of which follows.
Let
\begin{equation}\label{eq:ri}
\ri=Z_iS_i+(1-Z_i)\ppi=\begin{cases}
\ppi &Z_i=0\\
\sti &Z_i=1
\end{cases}
\end{equation}
In general, $(1-Z_i)\ppi=(1-Z_i)\ri$ and $Z_i\sti=Z_i\ri$, allowing $\ri$ to take the place of $\ppi$ and $S_i$ throughout $\tilde{\Psi}_i$.
Denote estimating equations based on $\ri$ as $\sum_i \psi_i=\bm{0}$.

@@

% If principal scores were known exactly, the M-estimation structure of $\heff0$ and $\heff1$ implies that their sampling variance could be estimated by the standard sandwich estimator for OLS \citep{stefanskiBoos}.
% This will rarely be the case.
%However,
If the principal scores are modeled as \eqref{eq:prinScoreMod}, and parameters $\bm{\alpha}$ are estimated by M-estimation, with estimating equations $\sum_i Z_i\Omega_i=\bm{0}$, then the estimating equations for the principal score model and principal effect estimation can be stacked as in \eqref{eq:stacked}.
Sandwich standard error estimates of the form \eqref{sandwich} based on $\left[Z_i\omega_i,\psi_i\right]$ will incorporate uncertainty from both principal score modeling and principal effect estimation conditional on principal scores.

@@

In any event, Assumption \ref{ass:rci} allows principal effects to be estimated with an OLS model including $\bxy$:
\begin{prop}{\textsc{geepers}}\label{prop:reg2}

@@

Potential and observed outcomes were generated as
\begin{equation}\label{eq:y-sim}
Y_i=\beta_1\sti+\beta_3Z_i\sti+(\gamma_1+\gamma_2\sti+\gamma_3Z_i)(x_{1i}+x_{2i})+\frac{1}{\sqrt{6}}x_{3i}+\epsilon_i%\\
\end{equation}
%   \begin{split}
%   \yci&=\beta_1\sti+(\gamma_1+\gamma_2\sti)(x_{1i}+x_{2i})+\frac{1}{\sqrt{6}}x_{3i}+\epsilon_i\\
%   \yti&=0.3\sti+(\gamma_1+\gamma_2\sti+\gamma_3Z_i)(x_{1i}+x_{2i})+\frac{1}{\sqrt{6}}x_{3i}+\epsilon_i\\
%   Y_i&=Z_i\yti+(1-Z_i)\yci
%   \end{split}
% \end{equation}
with $\beta_3=0.3-\beta_1$, $\EE[\epsilon_i]=0$, and $Var(\epsilon_i)=1/2$.
The distribution of $\epsilon_i$, $\beta_1$, and values of coefficients $\gamma_1$, $\gamma_2$, and $\gamma_3$ were manipulated factors in the study.
As described below, in most simulated scenarios, $\gamma_1=1/\sqrt{6}$ and $\gamma_2=\gamma_3=0$. In these cases there is no average treatment effect in the $\st=0$ stratum, and, since  $Var(\gamma_1(x_1+x_2+x_3))=1/2$ and $Var(\epsilon_i)=1/2$, covariates explained half of the variance of $Y$ within treatment group and principal stratum, and observed covariates $x_1$ and $x_2$ explained 1/3 of that variance.
These values are in line with estimates from outcome models in Section \ref{sec:fh2t} and are typical in education contexts \cite{hedgesHedberg}.

@@

\subsubsection{Data Analysis}
We estimated principal scores using logistic regression of observed $S$ on school fixed effects and a set of baseline variables $\bxs$.
Three criteria governed our choice of covariates and regression specifications for estimating principal scores and effects.
First, Proposition \ref{prop:reg2} requires consistent estimation of principal scores and (via Assumption \ref{ass:rci}) correct specification of the outcome model, that is, the model for $\EE[Y|\bxy,Z,\st]$.
Hence, it is essential to choose covariates and transformations $\bxs$ and $\bxy$ that lead to good model fit.

Second, Assumption \ref{ass:vps} requires principal scores to vary with at least some element $\bxs$. Moreover, the results of the simulation study suggest that when $\st$ can be predicted more precisely as a function of $\bxs$, \textsc{geeper}s principal effect estimators are more precise and more robust to model misspecification.
Those considerations militate in favor of choosing covariates $\bxs$ to optimize out-of-sample principal score prediction accuracy.

There is good reason to believe that analogous arguments apply to selection of outcome covariates $\bxy$.

The first two criteria motivated our choice of two strategies for choosing $\bxs$ and $\bxy$.
In our main model, we included every available covariate in the dataset and chose transformations of those covariates based on the results of residual and binned residual plots \citep{arm}.

Including all possible covariates maximizes the model's opportunities to detect covariate relationships that would predict $Y$ or $\st$.
However, this strategy can also lead to overfitting, which would decrease out-of-sample predictive accuracy.
In an alternative analysis, we chose a subset of covariates to include in $\bxs$ and $\bxy$ with a backward stepwise selection algorithm minimizing the AIC \citep{aic} of the principal score model and then modified the resulting models based on residual plots.
The model selection process is recorded in the GitHub repository cited above.

Lastly, Assumption \ref{ass:ci} implies that there is no interaction between $\bxs$ and $\st$ in the outcome model. Further, for the sake of statistical efficiency, we did not consider outcome models with interactions between covariates and $Z$.
\emph{A priori}, the student attribute that poses the greatest threat to these modeling choices is prior achievement, that is, scores on the pretest and the fifth-grade state test.
This is both because they measure a similar construct to the posttest, and because \cite{impactPaper} found a pretest-treatment interaction in a different treatment contrast.

Following that reasoning, we conducted a third parallel analysis excluding prior achievement covariates in $\bxs$ but including all covariates in $\bxy$.

@@

Outcome models for all analyses adjusted for all of the variables that were in the corresponding principal score models, in addition to measurements of students' prior knowledge; also, the models substituted class fixed effects for school fixed effects.\footnote{Why use a coarser set of fixed effects for the principal score model than the outcome regression? First, the principal score model was fit to a smaller sample (only students in the ``Instant'' condition), so some classrooms only included one student, which can lead to poor fit. Second, the asymptotics of logistic regression are more restricted than OLS, so logistic regression tends to perform poorly when the number of parameters increases with the sample size \citep{agresti}.}
As with principal score models, we checked model fit with residual plots and modified model specifications accordingly.

@@

\subsubsection{Results}

@@

\begin{figure}
  \centering
  \includegraphics{figure/compareMethods.pdf}
  \caption{Principal effect estimates with 95\% confidence or credible intervals using three different principal stratification methods, alongside ATE estimates and 95\% confidence intervals, for contrasts between the ASSISTments condition and each of the three other conditions in the study. The \textsc{geepers} estimates used the ``All Covariates'' principal score model.}
  \label{fig:compare}
\end{figure}

Figure \ref{fig:compare} shows principal effect estimates from \textsc{geepers} (using the All Covariates principal score model) compared to analogous estimates using Bayesian mixture modeling and \textsc{psw}, alongside estimates of the overall average treatment effects, labeled ``ATE.''

The \textsc{psw} model used the same principal score model as \textsc{geepers}, while the Bayesian model used the same principal score and outcome model specifications.

@@

\sloppy
Following \eqref{eq:stacked}, let $\blam(Z_i,S_i,\bxsi,\bxy_i,Y_i;\bm{\alpha},\bm{\beta})=\begin{pmatrix} Z_i\Omega(\bxsi,S_i;\bm{\alpha})\\ \Psi(\bxy_i,\bxsi,Y_i,Z_i,\ri(\bm{\alpha});\bm{\beta})\end{pmatrix}$, the stacked estimating equations of \eqref{eq:psMod} and \eqref{eq:regression}.
Going forward, for the sake of brevity, we will write $\blam_i(\bm{\alpha},\bm{\beta})=\blam(Z_i,S_i,\bxsi,\bxy_i,Y_i;\bm{\hat{\alpha}},\bm{\hat{\beta}})$, where dependence on the data for $i$ is captured in the subscript $i$, with similar meanings for $\Psi_i(\bm{\alpha},\bm{\beta})$ and $\omega_i(\bm{\alpha})$.
%Then let $\blamh_i=\blam_i(Z_i,S_i,\bxsi,\bxy_i,Y_i;\bm{\hat{\alpha}},\bm{\hat{\beta}})$ be the estimating equations evaluated at the estimated parameters, and let

%\begin{equation*}
%\dblam_i=\frac{\partial}{\partial [\alpha,\beta]'} \blam_i\Bigr|_{\substack{\bm{\alpha}=\bm{\hat{\alpha}}\\\bm{\beta}=\bm{\hat{\beta}}}}
%\end{equation*}
%the derivative matrix of $\blam$ evaluated at the estimated parameters.

@@

Following \citet[][p. 373]{carroll2006measurement}, we %separate the estimating equations $\Psi$ into $\phi$ and $\psi$, estimating equations for models \eqref{
can decompose the matrices into diagonal elements
\begin{equation*}
    \begin{split}
        A_{1,1}&=\sum_i \partial \Omega_i/\partial \bm{\alpha}|_{\bm{\alpha}=\bm{\hat{\alpha}}}\\
        A_{2,2}&=\sum_i\partial\Psi_i/\partial \bm{\beta}|_{\bm{\beta}=\bm{\hat{\beta}}}\\
        B_{1,1}&=\sum_i\Omega_i(\bm{\hat\alpha})\Omega_i(\bm{\hat\alpha})'\\
        B_{2,2}&=\sum_i \Psi_i(\bm{\hat\alpha},\bm{\hat\beta})\Psi_i(\bm{\hat\alpha},\bm{\hat\beta})'
    \end{split}
\end{equation*}
 that pertain to the parameter sets $\bm{\alpha}$ and $\bm{\beta}$ and the estimating equations for models \eqref{eq:psMod} and  \eqref{eq:regression}, respectively, and
 \begin{equation*}
     \begin{split}
         A_{21}&=\sum_i\partial\Psi_i/\partial \bm{\alpha}|_{\bm{\alpha}=\bm{\hat{\alpha}}}\\
         B_{12}=B_{21}'&=\sum_i \Omega_i(\bm{\hat{\alpha}})\Psi_i(\bm{\hat{\alpha}},\bm{\hat{\beta}})'
     \end{split}
 \end{equation*}
 which capture the dependence of model \eqref{eq:regression} on the parameters $\bm{\alpha}$ from \eqref{eq:psMod} and the covariance between the estimating equations of the two models.

 The sub-matrix $A_{12}=\sum_i \partial \Omega_i/\partial \bm{\beta}=0$, since \eqref{eq:psMod} does not depend on $\bm{\beta}$.

@@

\begin{center}

    \includegraphics[width=5.5in]{../simFigs/alphaAUC.pdf}

\end{center}
%\begin{table}
%\caption{Coverage of nominal 95\% confidence intervals}
\clearpage

\subsection{Full Empirical 95\% Interval Coverage Results}
The following table gives the empirical coverage of nominal 95\% intervals for \textsc{geepers} and mixture model principal effect estimates under varying data generating models.\\

\input{coverageTabAppendix.tex}

\clearpage
\subsection{Full RMSE Results}
The following table gives the root mean squared error (RMSE) for \textsc{geepers}, mixture model, and principal score weighting principal effect estimates under varying data generating models.\\

\input{rmseTabAppendix}

@@

\small
\input{tab1num}

@@

\small
%\input{writeUps/outcomeRegAppendix}
\input{psTab}

\input{regTab}
